Corpus Creation

### 2020-08-23 inst/analysis/corpus-analysis.Rmd
Prototype how could it be in docker

---
title: "Corpus Analysis"
output: 
  rmdformats::readthedown:
        code_folding: hide
        lightbox: true
        gallery: false
        df_print: paged
        toc_depth: 3
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
---

>>>>

---
title: "Corpus Analysis"
output: 
  rmdformats::readthedown:
        code_folding: hide
        lightbox: true
        gallery: false
        df_print: paged
        toc_depth: 3
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: /home/rstudio/evalR
  hostname: prl2
  port: 8788
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT", 
  "fs",
  "lubridate",
  "purrr",
  "readr",
  "rjson",
  "stringr",
  "tibble"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(echo = TRUE)

source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

create_tags(path(TAGS_DIR, "corpus.tex"), prefix="", default=TRUE)

# TODO
PACKAGE_COVERAGE_LOG <- "/home/rstudio/evalR/run/package-coverage/parallel.csv"
```

## Description

The aim of this notebook is to analyze the corpus of R packages.

## Loading data

This notebook requires the following data files:

- `r PACKAGE_COVERAGE_LOG` - package coverage log
- `r PACKAGE_COVERAGE` - package coverage

```{r load data - packages, include=FALSE}
coverage_log <- read_runr_log(PACKAGE_COVERAGE_LOG)
```

```{r}
failed_coverage <- filter(coverage_log, exitval != 0)
```

```{r}
failed_coverage %>%
  mutate(
    run_path=show_url(run_path)
  ) %>%
  my_datatable(escape=FALSE)
```

```{r load data - revdep, include=FALSE}
pkgs_revdeps_raw <- read_csv(
  CRAN_REVDEPS, 
  col_types=cols(
    package = col_character(),
    revdep = col_character()
  )
)
```

## Thresholds

```{r thresholds}
overview_table(
  r("threshold code coverage", percent(params$min_coverage)),
  r("threshold revdeps", params$min_revdeps),
  r("package size outier", params$package_size_outlier)
)
```

## CRAN Overview

### Table

```{r summary}
summary_corpus <- function(df, prefix) {
  w_coverage <- filter(df, coverage_expr > 0)
  w_revdeps <- filter(df, revdeps > 0)
  w_runnable <- filter(df, !is.na(runnable_time))

  overview_table(
    r(paste(prefix, "cran"), df),
    r(paste(prefix, "installed"), filter(df, installed)),
    r(paste(prefix, "loadable"), filter(df, loadable)),

    r(paste(prefix, "with runnable code"), filter(df, runnable_code > 0)),
    r(paste(prefix, "runnable code"), sum(df$runnable_code, na.rm=T)),
    r(paste(prefix, "examples code"), sum(df$runnable_code_examples, na.rm=T)),
    r(paste(prefix, "tests code"), sum(df$runnable_code_tests, na.rm=T)),
    r(paste(prefix, "vignettes code"), sum(df$runnable_code_vignettes , na.rm=T)),

    r(paste(prefix, "with coverage"), w_coverage),
    r(paste(prefix, "mean expr coverage"), percent(mean(w_coverage$coverage_expr))),
    r(paste(prefix, "median expr coverage"), percent(median(w_coverage$coverage_expr))),
    r(paste(prefix, "covered functions ratio"), ratio(sum(w_coverage$funs_with_coverage),sum(w_coverage$funs))),

    r(paste(prefix, "with R code"), filter(df, package_r_code > 0)),
    r(paste(prefix, "R code"), sum(df$package_r_code, na.rm=T)),
    r(paste(prefix, "with native code"), filter(df, package_native_code > 0)),
    r(paste(prefix, "native code"), sum(df$package_native_code, na.rm=T)),

    r(paste(prefix, "functions"), sum(df$funs, na.rm=T)),
    r(paste(prefix, "public functions"), sum(df$funs_public, na.rm=T)),
    r(paste(prefix, "private functions"), sum(df$funs_private, na.rm=T)),
    r(paste(prefix, "S three functions"), sum(df$funs_s3, na.rm=T)),
    r(paste(prefix, "S three classes"), sum(df$s3_classes, na.rm=T)),

    r(paste(prefix, "w revdes"), w_revdeps),
    r(paste(prefix, "revdes"), sum(w_revdeps$revdeps)),
    r(paste(prefix, "mean revdes"), mean(w_revdeps$revdeps)),
    r(paste(prefix, "median revdes"), median(w_revdeps$revdeps)),
    r(paste(prefix, "at least min revdep"), filter(w_revdeps, revdeps >= params$min_revdeps)),
    r(paste(prefix, "revdep runnable code"), sum(w_revdeps$revdeps_runnable_code, na.rm=T)),
    r(paste(prefix, "revdep mean code coverage"), percent(mean(w_revdeps$revdep_coverage_expr)/100)),
    r(paste(prefix, "revdep median code coverage"), percent(median(w_revdeps$revdep_coverage_expr)/100)),

    r(paste(prefix, "runtime"), as.duration(sum(w_runnable$runnable_time))),
    r(paste(prefix, "mean runtime"), as.duration(mean(w_runnable$runnable_time))),
    r(paste(prefix, "median runtime"), as.duration(median(w_runnable$runnable_time))),

    r(paste(prefix, "using roxygen"), sum(df$uses_roxygen, na.rm=T)),
  )
}

summary_corpus(pkgs, "All")
```

### Plots

```{r all considered packages}
pkgs_considered <- filter(pkgs, loadable, package_r_code > 0, coverage_expr > 0)
```

```{r plot all considered packages}
plot_all <- function(df) {
  coverage_mean <- mean(df$coverage_expr)
  coverage_median <- median(df$coverage_expr)
  revdeps_mean <- mean(df$revdeps)
  revdeps_median <- median(df$revdeps)

  df %>%
    mutate(
      outlier=package_code > 5e4,
      label=if_else(outlier, str_glue("{package}"), as.character(NA))
    ) %>%
    ggplot(
      aes(
        x=coverage_expr, 
        y=revdeps,
        label=label,
        size=package_code
      )
    ) +
    geom_point(alpha=.6) +
    geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +

    geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="black", size=0.2) +
    geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="black", size=0.2) +
    geom_vline(aes(xintercept=params$min_coverage),linetype="solid", color="red", size=0.4) +
    geom_hline(aes(yintercept=params$min_revdeps),linetype="solid", color="red", size=0.4) +

    scale_x_continuous(labels = scales::percent) + 
    scale_y_log10(labels = scales::comma) +
    scale_size(range = c(.1, 5), labels = scales::comma) +
      theme(
      legend.position=c(0.2, 0.85),
      legend.box="horizontal",
      legend.box.background = element_rect(fill="white", size=0.1)
    ) +
    labs(
      x="Code coverage", 
      y="Number of reverse dependencies (log)", 
      size="Code size",
      linetype=""
    ) +
    guides(
      size=guide_legend(order=1),
      color=guide_legend(order=2),
      linetype=F
    )
}

plot_all(pkgs_considered)
```

## Corpus overview

We only consider packages that have at least `r params$min_revdeps` reverse dependencies and `r params$min_coverage*100`% expression coverage.

```{r project corpus}
corpus <- 
  filter(
    pkgs, 
    loadable, 
    runnable_code > 0, 
    revdeps >= params$min_revdeps, 
    package_r_code > 0, 
    coverage_expr >= params$min_coverage
  )

write_lines(corpus$package, CORPUS_FILE)
write_csv(corpus, CORPUS_DETAILS)
```

### Reverse dependencies

```{r revdeps}
corpus_revdeps <- pkgs_revdeps_raw %>% semi_join(corpus, by="package")
write_csv(pkgs_revdeps_raw, CORPUS_REVDEPS)
write_lines(unique(corpus_revdeps$revdep), CORPUS_REVDEPS_UNIQUE)
```

### Table

```{r corpus summary}
summary_corpus(corpus, "corpus")

write_csv(corpus, CORPUS_DETAILS)
write_lines(corpus$package, CORPUS_FILE)
```

### Plot

```{r plot corpus}
plot_corpus <- function(df, code_outlier) {
  coverage_mean <- mean(df$coverage_expr)
  coverage_median <- median(df$coverage_expr)
  revdeps_mean <- mean(df$revdeps)
  revdeps_median <- median(df$revdeps)

  df %>%
    mutate(
      outlier=package_code > code_outlier,
      label=if_else(outlier, str_glue("{package}"), as.character(NA))
    ) %>%
    ggplot(
      aes(
        x=coverage_expr, 
        y=revdeps,
        label=label,
        size=package_code
      )
    ) +
    geom_point(alpha=.6) +
    geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +

    geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="red", size=0.2) +
    geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +

    scale_x_continuous(limits = c(params$min_coverage, 1), labels = scales::percent) + 
    scale_y_log10(labels = scales::comma) +
    scale_size(range = c(.1, 3), labels = scales::comma) +
      theme(
      legend.position=c(0.2, 0.85),
      legend.box="horizontal",
      legend.box.background = element_rect(fill="white", size=0.1)
    ) +
    labs(
      x="Code coverage", 
      y="Number of reverse dependencies (log)", 
      size="Code size",
      linetype=""
    ) +
    guides(
      size=guide_legend(order=1),
      color=guide_legend(order=2),
      linetype=F
    )
}

plot_corpus(corpus, params$package_size_outlier)
ggsave(CORPUS_PLOT)
```

### List of corpus packages

The following table is a detailed view of all the packages included in the corpus.

```{r list of packages}
datatable(
  corpus, 
  filter="top", 
  options=list(
    pageLength=50,
    autoWidth=TRUE
  )
)
```

### 2020-08-23 inst/analysis/corpus-creation.Rmd
Prototype how could it be in docker

@@

```{r package blacklist}
blacklist <- tribble(
  ~package,
  "ggm",      # requires bioc
  "shinyBS"   # runs shiny server as part of its tests
)
```

```{r get all available packages}
# TODO this should be installed.packages() once they are installed
#cran_packages_raw <- available.packages()
cran_packages_raw <- installed.packages()
cran_packages <- as_tibble(cran_packages_raw) %>%
  select(package=Package, version=Version) %>%
  anti_join(blacklist, by="package")
```

@@

```{r get last month most downloaded packages}
pkgs_top_downloaded <-
  cranlogs::cran_top_downloads("last-month", 100)
```

```{r compute missing packages from the most downloaded ones}
missing <- setdiff(pkgs_top_downloaded$package, pkgs_top_revdeps$package)
missing_revdeps <- filter(pkgs_revdeps, package %in% missing)
missing_revdeps %>%
  left_join(pkgs_top_downloaded, by="package") %>%
  datatable()
```

>>>>

```{r consts}
BASE_PACKAGES <- tribble(
  ~package,
  "base",
  "compiler",
  "datasets",
  "graphics",
  "grDevices",
  "grid",
  "methods",
  "parallel",
  "splines",
  "stats",
  "stats4",
  "tcltk",
  "tools",
  "utils"
)
```

@@

```{r package blacklist}
blacklist <- tribble(
  ~package,
  "ggm",       # requires bioc
  "future",    # does not work with OMP thread setting
  "shinyBS",   # runs shiny server as part of its tests
  "spatstat",  # takes too long
  "survival"   # takes too long
)
```

```{r get all available packages}
# TODO this should be installed.packages() once they are installed
#cran_packages_raw <- available.packages()
cran_packages_raw <- installed.packages()
cran_packages <- as_tibble(cran_packages_raw) %>%
  select(package=Package, version=Version) %>%
  anti_join(blacklist, by="package") %>%
  anti_join(BASE_PACKAGES, by="package")
```

@@

```{r get last month most downloaded packages, eval=F}
pkgs_top_downloaded <-
  cranlogs::cran_top_downloads("last-month", 100)
```

```{r compute missing packages from the most downloaded ones, eval=F}
missing <- setdiff(pkgs_top_downloaded$package, pkgs_top_revdeps$package)
missing_revdeps <- filter(pkgs_revdeps, package %in% missing)
missing_revdeps %>%
  left_join(pkgs_top_downloaded, by="package") %>%
  datatable()
```
